### 2024-06-03 thesis/appendix.tex
final

\section{Quantile regressor extensive comparison}\label{appendix:quantile_regressor_extensive_comparison}
This section contains extensive comparison between our kernel quantile regression and other state of the art quantile regressors. We benchmark it on popular machine learning datasets.
\subsection{Boston housing dataset}
The Boston housing dataset \href{https://www.kaggle.com/datasets/altavish/boston-housing-dataset}{https://www.kaggle.com/datasets/altavish/boston-housing-dataset} contains information about various attributes for suburbs in Boston.
There are 13 indipendent variables:
\begin{itemize}
\item CRIM: per capita crime rate by town
\item ZN: proportion of residential land zoned for lots over 25,000 sq.ft.\
\item INDUS: proportion of non-retail business acres per town
\item CHAS: Charles River dummy variable (1 if tract bounds river, 0 otherwise)
\item NOX: nitric oxides concentration (parts per 10 million)
\item RM: average number of rooms per dwelling
\item AGE: proportion of owner-occupied units built prior to 1940
\item DIS: weighted distances to five Boston employment centres
\item RAD: index of accessibility to radial highways
\item TAX: full-value property-tax rate per 10,000
\item PTRATIO: pupil-teacher ratio by town
\item B: $1000(Bk - 0.63)^2$ where Bk is the proportion of afroamericans by town
\item LSTAT: lower status of the population
\end{itemize}
The dependent variable is MEDV, that is the median value of owner occupied homes in \$1000's

@@

\subsection{Abalone dataset}
The abalone data \href{https://archive.ics.uci.edu/dataset/1/abalone}{https://archive.ics.uci.edu/dataset/1/abalone} consistst of measurements of abalone molluscs, the goal is predicting their age by building a model for estimating its number of rings; age is the number of rings plus 1.5.
The data has 8 attributes:
\begin{itemize}
    \item Sex: categorical variable either male, female or infant
    \item Length
    \item Diameter
    \item Height
    \item Whole height
    \item Shucked height
    \item Viscera weight
    \item Shell weight
\end{itemize}

@@

\subsection{Vehicle dataset}
This data contains info about used cars \href{https://www.kaggle.com/datasets/nehalbirla/vehicle-dataset-from-cardekho}{https://www.kaggle.com/datasets/ne-halbirla/vehicle-dataset-from-cardekho}, the predictors are:
\begin{itemize}
    \item Year
    \item Present\_price: ex showroom price
    \item Kms Driven
    \item Fuel type
    \item Seller type
    \item Transmission
    \item Owner: number of previous owners
\end{itemize}
The dependent variable is the selling price.

@@

\subsection{Randomized search}
Randomized search controls the number of steps by choosing smartly the hyperparameters to try in each iterations.
Let us say, there are 100 candidates and we set the number of iterations to 20 then the search will stop after the 20th iteration and return the best set among the hyperparameters observed.
The advantage is that it is much more quicker than gridsearch, but on the other hand its performance is worse than gridsearch.

@@

\subsection{Cross validation for time series data}
When data points are dependent on preceding values, we cannot use standard $K$-fold cross validation. The rationale is that $K$-fold will randomize the order of the data, thus it might happen to use future data to predict the past; we want to avoid such behaviour in a time series setting. 
When carrying out out any kind of cross validation, we must keep consistency in the way we evaluate our predictors during model selection and in the way we perform evaluation of the test data.
Hence, for time series crossvalidation we have the timeseries split procedure, see Figure \ref{fig:crossvalidationtimeseries} for a visualisation. The blue observations make up the training sets while the orange observations form the test sets.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries.png}
    \caption{Cross validation for one step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries}
\end{figure}
We then average all the observed losses in order to get an estimate for $Err$.
Notice, in the literature this procedure is sometimes also referred to evaluation on a rolling forecasting origin. This comes from the fact that at each iteration we push forward the origin of our forecast.
The same concept applies for multi step ahead forecasting, see Figure \ref{fig:crossvalidationtimeseries2}.
In predicting $\hat{y}_{N+m}$ we use as inputs $y_1, y_2, \dots, \hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$, where $\hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$ are one step ahead forecasts.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries2.png}
    \caption{Cross validation for $m$ step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries2}
\end{figure}

@@

\subsection{Data normalisation}\label{appendix:normalization}
With data normalization we transform the range of the feautures to a standard scale.
Such preprocessing step is essential when employing distance based algorithms like SVM or K-nearest neighbours. The rationale behind it is that by normalizing data we give an uniform weight to each feature in the learning process; in this way we do not favour larger scale features.
Examples of the most popular features scaling are:
\begin{itemize}
    \item Standard scaler: it computes the standard score $z$ of a sample $x$
    \\
    $z=\frac{x-\mu}{\sigma}$.
    \item MinMax scaler: it maps every data sample to the range $[0,1]$.
    \\
    $\frac{x-\min(X)}{\max(X)-\min(X)}$
    \item Robust scaler: it scales features using statistics that are robust to outliers.
    Essentially, it subtracts the median and then scales the data according to the interquantile range.
\end{itemize}
Many other data scaling algorithms exists, we refer the reader to a thorough comparison \citeW{scikitscalers}.
\\
We conclude this subsection with a custom example that motivates the need of feature scaling \citeW{scikitscale_example}. The idea is to compare the results of modelling the data with K-nearest neighbours on the unscaled data against the scaled data. The considered data is the wine recognition dataset \href{https://archive.ics.uci.edu/dataset/109/wine}{https://archive.ics.uci.edu/da-taset/109/wine}. The goal for this dataset is recognising from whose cultivator the wine comes based on two features with a completely different scale. The first feature has values in the [0,1000] range while the second feature has values contained in [1,10].
The unscaled and scaled version are compared in Figure \ref{fig:feature_scaler_example}.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/feature_scaler_example.png}
    \caption[Importance of feature scaling]{Importance of feature scaling \citeW{scikitscale_example}}
    \label{fig:feature_scaler_example}
\end{figure}
What we can conclude from the image is that, the model trained on scaled data is much better than the other.
On the left, we can see that distances between categories are impacted solely by the larger scale feature. Conversely, on the right, we have that the two features contribute equally in determining the neighbours.

\subsection{Data compression}
When the number of points $n$ is large, kernel methods suffer from high computational costs. 
Using kernel methods, we have that the storage cost is of the order $O(n^2)$ while the computational cost for finding the solution is of the order $O(n^3)$.
A significant speed up can be obtained thanks to low rank approximation.

\subsubsection{Nystrom decomposition}
The Nyström approximation involves storing a submatrix of the whole kernel matrix. Thus, storage and computational costs are reduced to $O(nm)$ and $O(nm^2)$ respectively.
Nyström works by selecting $m<n$ points, called representative points; it approximates $K$ as
\begin{equation}
    \tilde {K}=K_{n,m} K_{m,m}^{-1}K_{n,m}^\intercal
\end{equation}

@@

\newpage
\section{Source code}\label{src_code}
The whole code for the project is hosted on
\url{https://github.com/luca-pernigo/ThesisKernelMethods}\label{github_repo}.
\\
\begin{itemize}
    \item query/: folder containing Scopus data and scripts to generate bibliometric survey plots in Section \ref{literature_review}.
    \item thesis/: folder containing tex files for thesis pdf.
    \item beamer/: folder containing tex files for beamer pptx %pptx thesis defence.
    \item experiments: folder containing train, test script for each of the experiments carried out in the thesis alongside with the binary files of the pickled models.
    \item experiments/src/kernel\_quantile\_regression/kqr.py: file implementing our custom kernel quantile regression.
    \item experiments/Data: cleaned data used in our experimental studies.
    \item experiment/plots/: folder storing plots resulting from experiments.
    \item experiment/eda/: scripts for exploratory data analysis.
    \item experiments/point/: scripts for experiments point framework.
    \item experiments/train\_test/: scripts for training, testing and generating tables for experiments probabilistic framework.
    \item experiments/utils/: utility functions used in extracting, transforming and cleaning raw data.
    \item extradata/: folder containing data for experiments in section \ref{appendix:quantile_regressor_extensive_comparison}.
    \item requirements.txt: pip freeze of python packages used.
\end{itemize}

>>>>

\section{Quantile regressor extensive comparison}\label{appendix:quantile_regressor_extensive_comparison}
This section contains extensive comparison between our kernel quantile regression and other state of the art quantile regressors. We benchmark it on popular machine learning datasets.
\subsection{Boston housing dataset}
The Boston housing dataset \href{https://www.kaggle.com/datasets/altavish/boston-housing-dataset}{https://www.kaggle.com/datasets/altavish/boston-housing-dataset} contains information about various attributes for suburbs in Boston.
There are 13 indipendent variables
\begin{itemize}
\item CRIM: per capita crime rate by town
\item ZN: proportion of residential land zoned for lots over 25,000 sq.ft.\
\item INDUS: proportion of non-retail business acres per town
\item CHAS: Charles River dummy variable (1 if tract bounds river, 0 otherwise)
\item NOX: nitric oxides concentration (parts per 10 million)
\item RM: average number of rooms per dwelling
\item AGE: proportion of owner-occupied units built prior to 1940
\item DIS: weighted distances to five Boston employment centres
\item RAD: index of accessibility to radial highways
\item TAX: full-value property-tax rate per 10,000
\item PTRATIO: pupil-teacher ratio by town
\item B: $1000(Bk - 0.63)^2$ where Bk is the proportion of afroamericans by town
\item LSTAT: lower status of the population
\end{itemize}
The dependent variable is MEDV, that is the median value of owner occupied homes in \$1000's

@@

\subsection{Abalone dataset}
The abalone data \href{https://archive.ics.uci.edu/dataset/1/abalone}{https://archive.ics.uci.edu/dataset/1/abalone} consists of measurements of abalone molluscs, the goal is predicting their age by building a model for estimating its number of rings; age is the number of rings plus 1.5.
The data has 8 attributes
\begin{itemize}
    \item Sex: categorical variable either male, female or infant
    \item Length
    \item Diameter
    \item Height
    \item Whole height
    \item Shucked height
    \item Viscera weight
    \item Shell weight
\end{itemize}

@@

\subsection{Vehicle dataset}
This data contains info about used cars \href{https://www.kaggle.com/datasets/nehalbirla/vehicle-dataset-from-cardekho}{https://www.kaggle.com/datasets/ne-halbirla/vehicle-dataset-from-cardekho}, the predictors are
\begin{itemize}
    \item Year
    \item Present\_price: ex showroom price
    \item Kms Driven
    \item Fuel type
    \item Seller type
    \item Transmission
    \item Owner: number of previous owners
\end{itemize}
The dependent variable is the selling price.

@@

\subsection{Randomised search}
Randomised search controls the number of steps by choosing smartly the hyperparameters to try in each iteration.
Let us say, there are 100 candidates and we set the number of iterations to 20 then the search will stop after the $20^{th}$ iteration and return the best set among the hyperparameters observed.
The advantage is that it is much more quicker than gridsearch, but on the other hand its performance is worse than gridsearch.

@@

\subsection{Cross validation for time series data}
When data points are dependent on preceding values, we cannot use standard $K$-fold cross validation. The rationale is that $K$-fold will randomise the order of the data, thus it might happen to use future data to predict the past; we want to avoid such behaviour in a time series setting. 
When carrying out out any kind of cross validation, we must keep consistency in the way we evaluate our predictors during model selection and in the way we perform evaluation of the test data.
Hence, for time series crossvalidation we have the timeseries split procedure, see Figure \ref{fig:crossvalidationtimeseries} for a visualisation. The blue observations make up the training sets while the orange observations form the test sets.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries.png}
    \caption{Cross validation for one step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries}
\end{figure}
We then average all the observed losses in order to get an estimate for $Err$.
Notice, in the literature this procedure is sometimes also referred to evaluation on a rolling forecasting origin. This comes from the fact that at each iteration we push forward the origin of our forecast.
The same concept applies for multi step ahead forecasting, see Figure \ref{fig:crossvalidationtimeseries2}.
In predicting $\hat{y}_{N+m}$ we use as inputs $y_1, y_2, \dots, \hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$, where $\hat{y}_{N},\hat{y}_{N+1},\dots, \hat{y}_{N+m-1}$ are one step ahead forecasts.
\begin{figure}
    \includegraphics[width=\textwidth]{images/crossvalidationtimeseries2.png}
    \caption{Cross validation for $m$ step ahead timeseries data \cite{hyndman2018forecasting}}
    \label{fig:crossvalidationtimeseries2}
\end{figure}

@@

\subsection{Data normalisation}\label{appendix:normalization}
With data normalisation we transform the range of the features to a standard scale.
Such preprocessing step is essential when employing distance based algorithms like SVM or K-nearest neighbours. The rationale behind it is that by normalising data we give an uniform weight to each feature in the learning process; in this way we do not favour larger scale features.
Examples of the most popular features scaling are:
\begin{itemize}
    \item Standard scaler: it computes the standard score $z$ of a sample $x$
    \\
    $z=\frac{x-\mu}{\sigma}$.
    \item MinMax scaler: it maps every data sample to the range $[0,1]$.
    \\
    $\frac{x-\min(X)}{\max(X)-\min(X)}$
    \item Robust scaler: it scales features using statistics that are robust to outliers.
    Essentially, it subtracts the median and then scales the data according to the interquantile range.
\end{itemize}
Many other data scaling algorithms exists, we refer the reader to a thorough comparison \citeW{scikitscalers}.
\\
We conclude this subsection with a custom example that motivates the need of feature scaling \citeW{scikitscale_example}. The idea is to compare the results of modelling the data with K-nearest neighbours on the unscaled data against the scaled data. The considered data is the wine recognition dataset \href{https://archive.ics.uci.edu/dataset/109/wine}{https://archive.ics.uci.edu/da-taset/109/wine}. The goal for this dataset is recognising from whose cultivator the wine comes based on two features with a completely different scale. The first feature has values in the [0,1000] range while the second feature has values contained in [1,10].
The unscaled and scaled version are compared in Figure \ref{fig:feature_scaler_example}.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/feature_scaler_example.png}
    \caption[Importance of feature scaling]{Importance of feature scaling \citeW{scikitscale_example}}
    \label{fig:feature_scaler_example}
\end{figure}
What we can conclude from the image is that, the model trained on scaled data is much better than the other.
On the left, we can see that distances between categories are impacted solely by the larger scale feature. Conversely, on the right, we have that the two features contribute equally in determining the neighbours.

\subsection{Data compression}
When the number of points $n$ is large, kernel methods suffer from high computational costs. 
Using kernel methods, we have that the storage cost is of the order $O(n^2)$ while the computational cost for finding the solution is of the order $O(n^3)$.
A significant speed up can be obtained thanks to low rank approximations.

\subsubsection{Nyström decomposition}
The Nyström approximation involves storing a submatrix of the whole kernel matrix. Thus, storage and computational costs are reduced to $O(nm)$ and $O(nm^2)$ respectively.
Nyström works by selecting $m<n$ points, called representative points; it approximates $K$ as
\begin{equation}
    \tilde {K}=K_{n,m} K_{m,m}^{-1}K_{n,m}^\intercal
\end{equation}

@@

\newpage
\section{Source code}\label{src_code}
The whole code for the project is hosted on
\url{https://github.com/luca-pernigo/ThesisKernelMethods}\label{github_repo}.
\\
\begin{itemize}
    \item query/: folder containing Scopus data and scripts to generate bibliometric survey plots in Section \ref{literature_review}. Database queries have been taken from \cite{nowotarski}.
    \item thesis/: folder containing tex files for thesis pdf.
    \item beamer/: folder containing tex files for beamer pptx %pptx thesis defence.
    \item experiments: folder containing train, test script for each of the experiments carried out in the thesis alongside with the binary files of the pickled models.
    \item experiments/src/kernel\_quantile\_regression/kqr.py: file implementing our custom kernel quantile regression.
    \item experiments/Data: cleaned data used in our experimental studies.
    \item experiment/plots/: folder storing plots resulting from experiments.
    \item experiment/eda/: scripts for exploratory data analysis.
    \item experiments/point/: scripts for experiments point framework.
    \item experiments/train\_test/: scripts for training, testing and generating tables for experiments probabilistic framework.
    \item experiments/utils/: utility functions used in extracting, transforming and cleaning raw data.
    \item extradata/: folder containing data for experiments in section \ref{appendix:quantile_regressor_extensive_comparison}.
    \item requirements.txt: pip freeze of python packages used.
\end{itemize}
